COFFEE: an Optimizing Compiler for Finite Element Local Assembly

### 2014-06-23 TACO-paper/coffee.tex
Explain the three kernels

The actual complexity of a local assembly kernel depends on the finite element problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products involving basis functions. An example is provided in Figure~\ref{code:helmholtz}, which shows the assembly kernel for a Helmholtz problem using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving the Burgers equation, whose assembly code is given in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. This may also require the evaluation of constants in outer loops (called $F$ in the code) to act as scaling factors of arrays; trip counts can be larger (proportionally to the order of the method); and arrays may be block-sparse. In addition to a larger number of operations, more complex cases like the Burgers equation are characterized by high register pressure. The Helmholtz and Burgers equations, along with the Diffusion problem,

@@

Experiments were run on a single core of two Intel architectures, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and a Xeon Phi (5110P, running at 1.05Ghz in native mode, 32KB L1 cache and 512KB L2 cache). We have chosen these two architectures because of the differences in the number of logical registers and SIMD lanes, which can impact the effectiveness of the optimization strategy. The \texttt{icc 13.1} compiler was used. On the Sandy Bridge, the compilation flags used were \texttt{-O2} and \texttt{-xAVX} for auto-vectorization. On the Xeon Phi, optimization level \texttt{-O3} was used. Other optimization levels performed, in general, slightly worse. Our code transformations were evaluated in three real-world problems based on the following PDEs:
\begin{enumerate}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{enumerate}
The code was written in UFL (available at~\cite{ufl-code}) and then executed over real unstructured meshes through Firedrake. The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}. For Advection-Diffusion, the Diffusion equation, which uses the same differential operators as Helmholtz, is considered. In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$. Burgers is a non-linear problem employing differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (see Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are needed.

>>>>

The actual complexity of a local assembly kernel depends on the finite element problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products involving basis functions. An example is provided in Figure~\ref{code:helmholtz}, which shows the assembly kernel for a Helmholtz problem using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving the Burgers equation, whose assembly code is given in Figure~\ref{code:burgers}, the number of arrays involved in the computation of the element matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. This may also require the evaluation of constants in outer loops (called $F$ in the code) to act as scaling factors of arrays; trip counts can be larger (proportionally to the order of the method); and arrays may be block-sparse. In addition to a larger number of operations, more complex cases like the Burgers equation are characterized by high register pressure.

The Helmholtz and Burgers equations are exponents of a large class of problems and, as such, will constitute our benchmark, along with the Diffusion equation. We carefully motivate this choice in Section~\ref{sec:real-examples}.

@@

Experiments were run on a single core of two Intel architectures, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and a Xeon Phi (5110P, running at 1.05Ghz in native mode, 32KB L1 cache and 512KB L2 cache). We have chosen these two architectures because of the differences in the number of logical registers and SIMD lanes, which can impact the effectiveness of the optimization strategy. The \texttt{icc 13.1} compiler was used. On the Sandy Bridge, the compilation flags used were \texttt{-O2} and \texttt{-xAVX} for auto-vectorization. On the Xeon Phi, optimization level \texttt{-O3} was used. Other optimization levels performed, in general, slightly worse.

\subsection{The Helmholtz, Diffusion, and Burgers equations}
\label{sec:real-examples}
Our code transformations were evaluated in three real-world problems based on the following PDEs
\begin{enumerate}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{enumerate}
The three chosen benchmarks are completely real examples and comprise the core operators in some of the most frequently encountered finite element problems in scientific computing. The Helmholtz and Diffusion kernels are archetypal second order elliptic operators. They are complete and unsimplified examples of the operators which occur for diffusion and viscosity in fluids, and for imposing pressure in compressible fluids. For instance, they are both extensively used in climate modeling. Very similar operators, for which the same optimisations are expected to be equally effective, apply to elasticity problems, which are at the core of computational structural mechanics. The Burgers kernel is a typical example of a first order hyperbolic conservation law, which occurs in real applications whenever a quantity is transported by a fluid (the momentum itself, in our case). We chose this particular kernel since it applies to a vector-valued quantity, while the elliptic operators apply to scalar quantities. In this way, the operators we have selected as examples are characteristic of both the second and first order operators which dominate in fluids and solids simulations.

The code was written in UFL (available at~\cite{ufl-code}) and then executed over real unstructured meshes through Firedrake. The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}. For Advection-Diffusion, the Diffusion equation, which uses the same differential operators as Helmholtz, is considered. In the Diffusion kernel code, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$. Burgers is a non-linear problem employing differential operators different from those of Helmholtz and relying on vector-valued quantities, which has a major impact on the generated assembly code (see Figure~\ref{code:burgers}), where a larger number of basis function arrays ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are generated.
